DRAFT \\ From physical principles to classical and quantum \\ Hamiltonian and Lagrangian particle mechanics

### 2016-06-24 2015CompleteDerivation/completeDerivation.tex
Differentiable functions for classical mechanics

\begin{prop}\label{classical_vector space}
The state space $\mathcal{C}$ for a homogeneous classical (i.e. infinitesimally reducible) material is a vector space over $\mathbbm{R}$ isomorphic to a subspace of $C(\mathcal{S}) \equiv \{\rho:\mathcal{S} \rightarrow \mathbbm{R} \; | \; \rho$ is continuous$\}$, where $\mathcal{S}$ is the state space of an infinitesimal amount of material.
\end{prop}

@@

The ability to associate finite amount of material with sets of states allows us to define finite regions of $\mathcal{S}$ and compare them. Consider a region of position and momentum in phase space. If we are able to spread a finite amount of material into a non-infinitesimal uniform distribution, then we know we have a finite region. And the density will give us an indication of how big the region is: if we double the region, the density will halve. In other words: because we are describing densities over $\mathcal{S}$, we are able to give a unique measure\footnote{Here measure is intended in the mathematical sense (a real valued function of a $\sigma$-algebra) which is distinct from other connotations in physics or probability theory.} $\mu$ for the size of sets of $\mathcal{S}$. From the measure we get a metric $d\mu$ and our positive linear functionals become integrals, $\Lambda_U (\mathcal{c}) = \int_U \rho_{\mathcal{c}} d \mu$.

@@

\begin{prop}\label{integration}
	Let $\mathcal{S}$ be the state space for the particles of a homogeneous classical material. Let each state $\mathcal{s} \in \mathcal{S}$ be fully identified by a finite number of independent state variables. Then $\mathcal{S}$ is endowed with a Borel measure $\mu$. The state space $\mathcal{C}$ is isomorphic to the space of Lebesgue integrable continuous functions over $\mathcal{S}$. That is: $\mathcal{C} \cong C(\mathcal{S}) \cap  L^1(\mathcal{S}, \mu)$ where $L^1(\mathcal{S}, \mu) = \{ \rho : \mathcal{S} \rightarrow \mathbbm{R} \; | \; \int_{\mathcal{S}} |\rho| d\mu < \infty \}$.
\end{prop}

\begin{justification}
	We claim there exists a positive linear functional $\Lambda_U : \mathcal{C} \rightarrow \mathbbm{R}$ for each $U \subseteq \mathcal{S}$ such that $\Lambda_U = \Lambda_{\mathrm{int}(U)}$. Let $\Lambda_{U} : \mathcal{C} \rightarrow \mathbbm{R}$ be the functional that returns the amount of material in $\mathcal{c}$ compatible with the outcome associated with the open set $U \subseteq \mathcal{S}$.$\Lambda_U$ is well-defined. $U$ is associated with a physically distinguishable outcome. $\Lambda_U$ is linear. $\Lambda_U(a_1 \mathcal{c}_1 + a_2 \mathcal{c}_2) = a_1 \Lambda_U(\mathcal{c}_1) + a_2 \Lambda_U(\mathcal{c}_2)$ as it has to be consistent with the operations of composing states and increasing/decreasing the amount of material by a factor. $\Lambda_U$ is positive. If the value of the distribution for each particle state is positive then the total amount of material is positive. Let $U \subseteq \mathcal{S}$ not necessarily open. Define $\Lambda_U$ as $\Lambda_{\mathrm{int}(U)}$.

@@

	We claim $\mathcal{C} \cong C(\mathcal{S}) \cap  L^1(\mathcal{S}, \mu)$. As per \ref{classical_vector space}, $\mathcal{C} \cong G \subseteq C(\mathcal{S})$ as a vector space over $\mathbbm{R}$. As $|\Lambda_{U}(\mathcal{c})| < \infty \; \forall \mathcal{c} \in \mathcal{C} \; \forall U \subseteq \mathcal{S}$, then $G \subseteq \{ \rho : S \rightarrow \mathbbm{R} \; | \;\; |\int_{U} \rho d\mu| < \infty \; \forall U \subseteq S\} = L^1(\mathcal{S}, \mu)$ by \ref{everywhere_integrable_is_lebesgue_integrable}.  $G \subseteq C(\mathcal{S}) \cap  L^1(\mathcal{S}, \mu)$. Let $\rho_\mathcal{c}$ be the distribution associated with the state $\mathcal{c}$. Let $\rho \in C(\mathcal{S}) \cap  L^1(\mathcal{S}, \mu)$ be a Lebesgue integrable continuous function. If $\rho_\mathcal{c} \neq \rho$ then $\exists U \in \mathcal{S} \; | \; \int_{U} \rho_\mathcal{c} d \mu \neq \int_{U} \rho d \mu$. There exists an outcome in $\mathcal{S}$ that can physically distinguish the two distributions. There must be a state $\mathcal{c}_1 \in \mathcal{C} \; | \; \varrho(\mathcal{c}_1)=\rho$. $\varrho : \mathcal{C} \rightarrow C(\mathcal{S}) \cap  L^1(\mathcal{S}, \mu)$ is surjective. $\varrho$ is an isomorphism between $\mathcal{C}$ and $C(\mathcal{S}) \cap  L^1(\mathcal{S}, \mu)$ as vector spaces.
\end{justification}

As we identified the set of Lebesque integrable continuous functions $C(\mathcal{S}) \cap  L^1(\mathcal{S}, \mu)$ as the set of distributions that are physically meaningful, we can better understand why other commonly used function spaces do not fit the bill. Some are not restrictive enough. The set of Lebesgue integrable functions $L^1(\mathcal{S})$ includes discontinuous functions that for \ref{continuous_map} are unphysical. The set of continuous functions $C(\mathcal{S})$ and the set of continuous functions that vanish at infinity $C_0(\mathcal{S})$ include functions whose integral is infinite, which would represent infinite amount of material. Some are too restrictive. The set of continuous functions with compact support $C_C(\mathcal{S})$ would exclude distributions, such as Gaussians, that span over the whole range of states. Such distributions are physical: if we take a finite volume of an ideal gas at equilibrium, the momentum distribution spans over the whole range. Schwartz space $S(\mathcal{S})$ excludes functions that are not infinitely smooth which we don't have a general physical justification for.\footnote{One can, though, make the argument that given any distribution $\rho$ one can find an infinitely smooth $\rho_{sm}$ such that the difference in description is small compared to the error already introduced by assuming the system to be homogeneous and infinitely reducible. }

@@

We now turn our attention to the metric $d\mu$ which we need to properly identify. It turns out that before we can do so, we need to better characterize $\mathcal{S}$.

Under an arbitrary change of state variables, the density has to remain the same, which means that the Jacobian of the transformation must exist and must be unitary. For the Jacobian to exist the transformation between state variables has to be differentiable. Therefore a manifold that guarantees densities to be well defined is one that guarantees that state variable transformations are differentiable: a differentiable manifold. $\mathcal{S}$ has a differentiable structure in the sense that the distributions are only defined on a set of set state variables that are linked by differential transformations. In the same way that discontinuous changes of state variable can be detected because they are incompatible with the topology, non differentiable changes of state variable can be detected because they are incompatible with state-variable-independent densities.

If $\rho_\mathcal{c}$ is to remain unchanged and the Jacobian to be unitary we cannot simply change one state variable as we please. If we change one at least another has to change in some coordinated way such that the Jacobian of the total transformation is unitary. This means that we cannot change physical units of all the variables as we please: they are part of a unit system. As we know, changing units of position will also change units of momentum. Let $\mathcal{Q}$ be the manifold charted by the minimum set of state variables that define such system of units. An element of $\mathcal{S}$ will be a geometrical object in such space. For example, it will be a point, a distance, two points, the ratio between two distances, the inverse of a distance: something that is expressible in the unit defined. As particle states represent configurations of infinitesimal parts, they will be local objects, defined around a point in $\mathcal{Q}$. As we saw before, they cannot be points themselves as the density $\rho(q')_\mathcal{c}= \rho(q)_\mathcal{c} dq/dq'$ is not invariant. They cannot be a vectors $v(q) e_q$, as the density $\rho_\mathcal{c}(q', v')= \rho(q, v)_\mathcal{c} (dq/dq')^2$ is not invariant. They can, however, be covectors $k(q) dq$ as $\rho_\mathcal{c}(q', k')= \rho_\mathcal{c}(q, k) (dq/dq') (dq'/dq) = \rho_\mathcal{c}(q, k)$.\footnote{The components $k$ are the classical analogue of the wave numbers.} In other words: a covector is the simplest local geometrical object that allows coordinate independent densities. Physically, the component $k_i$ of a covector, for example the wave number, can be expressed with a unit that is the inverse of the corresponding $q^i$. The product $\Delta q^i \Delta k_i$ is therefore a dimensionless quantity, invariant under unit transformation.  We can identify $\mathcal{S}$ as the cotangent bundle $T^*\mathcal{Q}$, the space of all possible covectors.

@@

We should note that we are starting to be in a position similar to the one discussed in \ref{discrete_state_metric} for discrete states. We have both a topological space and a metric that allows us to count states and possibilities. We recovered the metric starting from the idea of an infinitesimally reducible system. That led to a state space $\mathcal{S}$ for the infinitesimal parts, on which we must be able to define state-variable-independent distributions, which in turn gave us degrees of freedom made by pairs of conjugate variables and the symplectic metric typical of classical phase space. In short: being able to measure the amount of material is what allows us to measure the cardinality of states.

>>>>

\begin{prop}\label{classical_vector space}
The state space $\mathcal{C}$ for a homogeneous classical (i.e. infinitesimally reducible) material is a vector space over $\mathbbm{R}$ isomorphic to a subspace of the space of continuous functions. That is $\mathcal{C} \cong G \subseteq C(\mathcal{S}) \equiv \{\rho:\mathcal{S} \rightarrow \mathbbm{R} \; | \; \rho$ is continuous$\}$, where $\mathcal{S}$ is the state space of an infinitesimal amount of material.
\end{prop}

@@

The ability to associate finite amounts of material with sets of states allows us to define finite regions of $\mathcal{S}$ and compare them. Consider a region of position and momentum in phase space. If we are able to spread a finite amount of material into a non-infinitesimal uniform distribution, then we know we have a finite region. And the density will give us an indication of how big the region is: if we double the region, the density will halve. In other words: because we are describing densities over $\mathcal{S}$, we are able to give a unique measure\footnote{Here measure is intended in the mathematical sense (a real valued function of a $\sigma$-algebra) which is distinct from other connotations in physics or probability theory.} $\mu$ for the size of sets of $\mathcal{S}$. From the measure we get a metric $d\mu$ and our positive linear functionals become integrals, $\Lambda_U (\mathcal{c}) = \int_U \rho_{\mathcal{c}} d \mu$.

@@

\begin{prop}\label{integration}
	Let $\mathcal{S}$ be the state space for the particles of a homogeneous classical material. Let each state $\mathcal{s} \in \mathcal{S}$ be fully identified by a finite number of independent state variables. Then $\mathcal{S}$ is endowed with a Borel measure $\mu$. The state space $\mathcal{C}$ for a homogeneous classical material is isomorphic to a subspace of the space of Lebesgue integrable functions. That is $\mathcal{C} \cong G \subseteq L^1(\mathcal{S}, \mu) = \{ \rho : \mathcal{S} \rightarrow \mathbbm{R} \; | \; \int_{\mathcal{S}} |\rho| d\mu < \infty \}$.
\end{prop}

\begin{justification}
	We claim there exists a positive linear functional $\Lambda_U : \mathcal{C} \rightarrow \mathbbm{R}$ for each $U \subseteq \mathcal{S}$ such that $\Lambda_U = \Lambda_{\mathrm{int}(U)}$. Let $\Lambda_{U} : \mathcal{C} \rightarrow \mathbbm{R}$ be the functional that returns the amount of material in $\mathcal{c}$ compatible with the outcome associated with the open set $U \subseteq \mathcal{S}$. $\Lambda_U$ is well-defined: $U$ is associated with a physically distinguishable outcome. $\Lambda_U$ is linear: $\Lambda_U(a_1 \mathcal{c}_1 + a_2 \mathcal{c}_2) = a_1 \Lambda_U(\mathcal{c}_1) + a_2 \Lambda_U(\mathcal{c}_2)$ as it has to be consistent with the operations of composing states and increasing/decreasing the amount of material by a factor. $\Lambda_U$ is positive: if the value of the distribution for each particle state is positive then the total amount of material is positive. Let $U \subseteq \mathcal{S}$ not necessarily open. Define $\Lambda_U$ as $\Lambda_{\mathrm{int}(U)}$.

@@

	We claim $\mathcal{C}$ is isomorphic to a subspace of $L^1(\mathcal{S}, \mu)$ as a vector space over $\mathbbm{R}$. $\forall \mathcal{c} \in \mathcal{C} \exists ! \rho_{\mathcal{c}}:\mathcal{S} \rightarrow \mathbbm{R}$ returning the amount of material for each state $\mathcal{s} \in \mathcal{S}$. As $|\Lambda_{U}(\mathcal{c})| < \infty \; \forall \mathcal{c} \in \mathcal{C} \; \forall U \subseteq \mathcal{S}$, then $\rho_{\mathcal{c}} \in \mathcal{L}(S,\mu) \equiv \{ \rho : S \rightarrow \mathbbm{R} \; | \;\; |\int_{U} \rho d\mu| < \infty \; \forall U \subseteq S\}$. $\mathcal{L}(S,\mu) = L^1(\mathcal{S}, \mu)$ by \ref{everywhere_integrable_is_lebesgue_integrable}. Let $\varrho : \mathcal{C} \rightarrow L^1(\mathcal{S}, \mu)$ such that $\varrho(\mathcal{c}) \mapsto \rho_\mathcal{c}$. $\varrho$ is a homomorphism as we saw in the proof for \ref{classical_vector space}. $\mathcal{C}$ is isomorphic to $\varrho(\mathcal{C}) \subseteq L^1(\mathcal{S}, \mu)$ as a vector space.
\end{justification}

The first order of business is to guarantee that the density remains defined under an arbitrary change of state variables. This means the Jacobian of the transformation must exist. For the Jacobian to exist the transformation between state variables has to be differentiable. Therefore a manifold that guarantees densities to be well defined is one that guarantees that state variable transformations are differentiable: a differentiable manifold. $\mathcal{S}$ has a differentiable structure in the sense that the distributions and the volume element $d\mu$ are only defined on a set of state variables that are linked by differential transformations. In the same way that discontinuous changes of state variable can be detected because they are incompatible with the topology, non differentiable changes of state variable can be detected because they are incompatible with state-variable-independent densities.

Moreover, the density itself has to be differentiable. The distribution $\rho_\mathcal{c}$ is a real value function of the state. As such, at least locally, we can use it as a state variable $q^1=\rho_\mathcal{c}$. As a state variable it identifies the states by the density of the material associated with them. This is more physically meaningful as one may first suspect, since placing physical markers is a common way to define the references for a coordinate system, which in the end is placing material at particular states. In a coordinate system where $q^1=\rho_\mathcal{c}$, the function is clearly differential as $\partial_1 \rho_\mathcal{c} = 1$ and $\partial_i \rho_\mathcal{c} = 0$ for $i \neq 1$. As we change state variables, the transformation is differentiable, therefore all the derivatives $\partial_{i'} q^1 = \partial_{i'} \rho_\mathcal{c}$ exist. Therefore the density itself is differentiable.

\begin{prop}\label{differentiable_manifold}
	The state space $\mathcal{S}$ for the particles of a homogeneous classical material identified by a finite number of continuous independent state variables is a differentiable manifold. The state space $\mathcal{C}$ for a homogeneous classical material is isomorphic to the space of Lebesgue integrable differentiable functions. That is $\mathcal{C} \cong C^1(\mathcal{S}) \cap L^1(\mathcal{S}, \mu)$, where $C^1(\mathcal{S}) \equiv \{\rho:\mathcal{S} \rightarrow \mathbbm{R} \; | \; \rho$ is differentiable$\}$.
\end{prop}
\begin{justification}
	TODO
%Let $\rho_\mathcal{c}$ be the distribution associated with the state $\mathcal{c}$. Let $\rho \in C(\mathcal{S}) \cap  L^1(\mathcal{S}, \mu)$ be a Lebesgue integrable continuous function. If $\rho_\mathcal{c} \neq \rho$ then $\exists U \in \mathcal{S} \; | \; \int_{U} \rho_\mathcal{c} d \mu \neq \int_{U} \rho d \mu$. There exists an outcome in $\mathcal{S}$ that can physically distinguish the two distributions. There must be a state $\mathcal{c}_1 \in \mathcal{C} \; | \; \varrho(\mathcal{c}_1)=\rho$. $\varrho : \mathcal{C} \rightarrow C(\mathcal{S}) \cap  L^1(\mathcal{S}, \mu)$ is surjective. $\varrho$ is an isomorphism between $\mathcal{C}$ and $C(\mathcal{S}) \cap  L^1(\mathcal{S}, \mu)$ as vector spaces.
\end{justification}

As we identified the space of Lebesque integrable differentiable functions $C^1(\mathcal{S}) \cap  L^1(\mathcal{S}, \mu)$ as the space of distributions that are physically meaningful, we can better understand why other commonly used function spaces do not fit the bill. Some are not restrictive enough. The set of Lebesgue integrable functions $L^1(\mathcal{S})$ includes discontinuous functions that for \ref{continuous_map} are unphysical. The set of continuous functions $C(\mathcal{S})$ and the set of continuous functions that vanish at infinity $C_0(\mathcal{S})$ not only include non-differentiable functions, but also include functions whose integral is infinite, which would represent infinite amount of material. Such functions are also present in the space of differentiable functions $C^1(\mathcal{S})$. Some definitions are too restrictive. Requiring compact support (i.e.the function is different from zero only on a finite region) would exclude distributions, such as Gaussians, that span over the whole range of states. Such distributions are physical: if we take a finite volume of an ideal gas at equilibrium, the momentum distribution spans over the whole range. Schwartz space $S(\mathcal{S})$ excludes functions that are not infinitely smooth which we don't have a general physical justification for.\footnote{One can, though, make the argument that given any distribution $\rho$ one can find an infinitely smooth $\rho_{sm}$ such that the difference in description is small compared to the error already introduced by assuming the system to be homogeneous and infinitely reducible. }

@@

Now that we are guaranteed that the density is properly defined for all state variables, we have to make sure that is invariant. The distribution should only depend on the states, and not the particular choice of state variables.

If $\rho_\mathcal{c}$ is to remain invariant under state variable changes, not only the Jacobian has to exist but must be unitary. This means we cannot simply change one state variable as we please. If we change one at least another has to change in some coordinated way such that the Jacobian of the total transformation is unitary. This means that we cannot change physical units of all the variables as we like: they are part of a unit system. As we know, changing units of position will also change units of momentum. Let $\mathcal{Q}$ be the manifold charted by the minimum set of state variables that define such system of units. An element of $\mathcal{S}$ will be a geometrical object in such space. For example, it will be a point, a distance, two points, the ratio between two distances, the inverse of a distance: something that is expressible in the unit defined. As particle states represent configurations of infinitesimal parts, they will be local objects, defined around a point in $\mathcal{Q}$. As we saw before, they cannot be points themselves as the density $\rho(q')_\mathcal{c}= \rho(q)_\mathcal{c} dq/dq'$ is not invariant. They cannot be a vectors $v(q) e_q$, as the density $\rho_\mathcal{c}(q', v')= \rho(q, v)_\mathcal{c} (dq/dq')^2$ is not invariant. They can, however, be covectors $k(q) dq$ as $\rho_\mathcal{c}(q', k')= \rho_\mathcal{c}(q, k) (dq/dq') (dq'/dq) = \rho_\mathcal{c}(q, k)$.\footnote{The components $k$ are the classical analogue of the wave numbers.} In other words: a covector is the simplest local geometrical object that allows coordinate independent densities. Physically, the component $k_i$ of a covector, for example the wave number, can be expressed with a unit that is the inverse of the corresponding $q^i$. The product $\Delta q^i \Delta k_i$ is therefore a dimensionless quantity, invariant under unit transformation.  We can identify $\mathcal{S}$ as the cotangent bundle $T^*\mathcal{Q}$, the space of all possible covectors.

@@

It should be clear by now that the case of discrete topology is qualitatively different from the standard topology for $\mathbbm{R}^n$. The notion that the continuous case is a limit of the discrete case, that it's ``like the discrete but with more points", leads in this case to erroneous intuition. The question is: can we identify a single state? Can we have an outcome associated with only one element? The answer is yes in the discrete topology and is no in the continuous case. The consequence is that when we define the measure $\mu$ for the cardinality of continuous states, we assign a finite value to a finite region, and we assign zero to a single state. That is what gave us integration, densities and, ultimately, conjugate variables. If we were to use the discrete topology on $\mathbbm{R}^n$, if we were to assume we can identify single states, then we would assign measure one to each state and infinity to a finite region. We would not have integration, just a simple sum. Our distribution would not be a density and there would be no justification for conjugate variables. A finite distribution could only distribute finite amount of material in a finite number of states.

In other words: this is not a case were the difference at the small scale has a small difference in the large scale. The two cases are radically different. We can tell them apart. We should take the use of densities and conjugate quantities as empirical evidence that the quantities like space and time are not discrete. Or more precisely: that the processes we use to distinguish those quantities cannot identify single instances, as that is all we can say experimentally. The quantum case does not change this, as the use of densities (in the form of the wave function) and conjugate quantities are even more prominent.

Finally, we should note that we in a position similar to the one discussed in \ref{discrete_state_metric} for discrete states. We have both a topological space and a metric that allows us to count states and possibilities. We recovered the metric starting from the idea of an infinitesimally reducible system. That led to a state space $\mathcal{S}$ for the infinitesimal parts, on which we must be able to define state-variable-independent distributions, which in turn gave us degrees of freedom made by pairs of conjugate variables and the symplectic metric typical of classical phase space. In short: being able to measure the amount of material is what allows us to measure the cardinality of states.
